---
title: "Alternatives to RStudio"
subtitle: "Workflow organisation with R"
date: 17/12/2020
author: |
    | Nikita Gusarov
    | \scriptsize GAEL (UGA) - G-SCOP (Grenoble INP)
output: 
    beamer_presentation:
        theme: "Montpellier"
        colortheme: "beaver"
        toc: FALSE
        df_print: "kable"
        slide_level: 2
        fig_width: 4.2
        fig_height: 2.5
        fig_caption: TRUE
        includes:
            in_header: ../auxilary/header.tex
classoption: "aspectratio=169"
# header-includes: 
fontsize: 11pt 
---

# Introduction

## Workflow

```{r include = FALSE}
library(tidyverse)
```

<!-- 
Insert here RStudio, Atom, Jupyter, VSCode, ...

Speak about: the aim of this presentation, objectives, etc

- Present the problematics of the IDE choice
- Remind about tastes
- Remond about the absence of the absolute choice
-->

```{r eval = TRUE, echo = FALSE, fig.align = 'center'}
data <- data.frame(
    Editor = c(
        "Visual Studio Code", "Visual Studio", "Notepad++", 
        "IntelliJ", "Vim", "Sublime Text", "Android Studio", 
        "Eclipse", "PyCharm", "Atom", "IPython / Jupyter", "Xcode", 
        "PHPStorm", "NetBeans", "Emacs", "RStudio", "RubyMine", 
        "TextMate", "Coda", "Komodo", "Zend", "Light Table"
    ),
    Popularity = c(
        50.7, 31.5, 30.5, 25.4, 25.4, 23.4, 16.9, 14.4, 13.4, 13.3, 
        9.5, 9.4, 7.6, 5.9, 4.5, 3.4, 1.4, 0.9, 0.7, 0.4, 0.4, 0.2 
    )
)

data %>% 
    ggplot(aes(x = reorder(Editor, - Popularity), Popularity)) + 
        geom_bar(stat = "identity") +
        theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1)) + 
        labs(x = "Editor")
```

## Workflow types

- Text editor
- Notebook
- Integrated Development Environment (IDE)

## 

\vspace*{-10.6mm}\hspace*{-\beamerleftmargin}\hspace*{-1.5mm}
\fullframegraphic{"../captures/Capture dՎcran (0)"}

## 

\vspace*{-10.6mm}\hspace*{-\beamerleftmargin}\hspace*{-1.5mm}
\fullframegraphic{"../captures/Capture dՎcran (1)"}

## 

\vspace*{-10.6mm}\hspace*{-\beamerleftmargin}\hspace*{-1.5mm}
\fullframegraphic{"../captures/Capture dՎcran (3)"}



# What is RStudio ?

##

\vspace*{-10.6mm}\hspace*{-\beamerleftmargin}\hspace*{-1.5mm}
\fullframegraphic{"../captures/Capture dՎcran (2)"}

## "An IDE that was built for R"

- Syntax highlighting, code completion, and smart indentation
- Execute R code directly from the source editor
- Quickly jump to function definitions

## "Helps bring your workflow together"

- Integrated R help and documentation
- Easily manage multiple working directories using projects
- Workspace browser and data viewer

## "Powerful authoring and debugging"

- Interactive debugger to diagnose and fix errors quickly
- Extensive package development tools
- Authoring with Sweave and R Markdown

## And more ...

- Runs on most desktops or on a server and accessed over the web
- Integrates the tools you use with R into a single environment
- Includes powerful coding tools designed to enhance your productivity
- Enables rapid navigation to files and functions

## And more ...

- Makes it easy to start new or find existing projects
- Has integrated support for Git and Subversion
- Supports authoring HTML, PDF, Word Documents, and slide shows
- Supports interactive graphics with Shiny and ggvis

## 

\vspace*{-10.6mm}\hspace*{-\beamerleftmargin}\hspace*{-1.5mm}
\fullframegraphic{"../captures/Capture dՎcran (3)"}



# What RStudio is not ...

## Not a multi-language DE 

\vspace*{-10.6mm}\hspace*{-\beamerleftmargin}\hspace*{-1.5mm}
\fullframegraphic{"../captures/Capture dՎcran (12)"}

<!-- 
Add here some information about the alternative languages
-->



# What about alternatives ?



# Jupyter

## 

\vspace*{-10.6mm}\hspace*{-\beamerleftmargin}\hspace*{-1.5mm}
\fullframegraphic{"../captures/Capture dՎcran (15)"}

## Key advertising factors

<!--
The notebook extends the console-based approach to interactive computing in a qualitatively new direction, providing a web-based application suitable for capturing the whole computation process: developing, documenting, and executing code, as well as communicating the results. 
The Jupyter notebook combines two components:
-->

* A web application 

<!--
a browser-based tool for interactive authoring of documents which combine explanatory text, mathematics, computations and their rich media output.
-->

* Notebook documents

<!--
a representation of all content visible in the web application, including inputs and outputs of the computations, explanatory text, mathematics, images, and rich media representations of objects.
-->

## Features

- In-browser editing for code
- Automatic syntax highlighting, indentation, and tab completion
- The ability to execute code from the browser
- Results of computations is attached to the code which generated them

## Features 

- Displaying the result of computation using rich media representations (HTML, LaTeX, PNG, SVG, etc) 
- In-browser editing for rich text using the Markdown markup language
- Commentary for the code is not limited to plain text
- Include mathematical notation within markdown cells using LaTeX, and rendered natively by MathJax.

<!-- Add here some screenshots / pics for clarity -->

## 

\vspace*{-10.6mm}\hspace*{-\beamerleftmargin}\hspace*{-1.5mm}
\fullframegraphic{"../captures/Capture dՎcran (1)"}

## Drawbacks 

## Version control

<!--
Aff screenshot

Version control of files with tools such as git is important as it both maintains an explorable database of changes to the code files and also improves collaboration by using a centralized server (e.g. GitHub) where anyone with access to the repository can pull and push changes to the code. In the data science world, large startups such as Stripe and Airbnb have seen a lot of success with this approach.

RStudio incidentally has a native git client for tracking and committing changes to a .Rmd file, which is easy since .Rmd files are effectively plain text files where you can see differences between versions at a per-line level. (You may not want to store the changes to the generated .nb.html Notebook since they will be large and redundant to the changes made in the corresponding .Rmd; I recommend adding a *.nb.html rule to a .gitignore file during analysis).
-->

The .ipynb Jupyter Notebook files are blobs of JSON that also store cell output as well as metadata. 

<!--
This can cause the git database to balloon and makes reading per-line diffs hard if not impossible.

On Hacker News, the version control issues in Jupyter are a common complaint, however a Jupyter developer noted of a possibility of working with RStudio on solving this issue.
-->

##

\vspace*{-10.6mm}\hspace*{-\beamerleftmargin}\hspace*{-1.5mm}
\fullframegraphic{"../captures/Capture dՎcran (14)"}

## Inline Code Rendering

<!--
Add image

A common practice in Jupyter Notebooks is to print common values as a part of a write-up or testing statistical code. 
-->

In Jupyter Notebooks, it is impossible to use the inline expressions without additional markdown modules.

<!--
In R Notebooks, you can skip a step by calling such print statements in-line in the Markdown text, which will then be rendered with the Notebook. This also avoids hard-coding such numbers in
-->



# VS Code 

## 

\vspace*{-10.6mm}\hspace*{-\beamerleftmargin}\hspace*{-1.5mm}
\fullframegraphic{"../captures/Capture dՎcran (4)"}

## VS Code selling features

- Simplicity of a source code editor 
- Powerful developer tooling (IntelliSense, code completion and debugging)

## Available for macOS, Linux, and Windows

Visual Studio Code supports macOS, Linux, and Windows - so you can hit the ground running, no matter the platform.

## "Edit, build, and debug with ease"

- Lightning fast source code editor
- Support for hundreds of languages
- Intuitive keyboard shortcuts, easy customization and community-contributed mappings
- Interactive debugger
- Build and scripting tools to perform common tasks
- Support for Git so you can work with source control without leaving the editor including viewing pending changes diffs

## "Make it your own"

- Customization through extensions
- Open-source project

## 

\vspace*{-10.6mm}\hspace*{-\beamerleftmargin}\hspace*{-1.5mm}
\fullframegraphic{"../captures/Capture dՎcran (5)"}

## 

\vspace*{-10.6mm}\hspace*{-\beamerleftmargin}\hspace*{-1.5mm}
\fullframegraphic{"../captures/Capture dՎcran (6)"}

## 

\vspace*{-10.6mm}\hspace*{-\beamerleftmargin}\hspace*{-1.5mm}
\fullframegraphic{"../captures/Capture dՎcran (8)"}

## 

\vspace*{-10.6mm}\hspace*{-\beamerleftmargin}\hspace*{-1.5mm}
\fullframegraphic{"../captures/Capture dՎcran (13)"}

## References 

\small 

- [StackOverflow survey](https://insights.stackoverflow.com/survey/2019#technology-_-most-popular-development-environments)
- [Visual Studio Code](https://code.visualstudio.com/docs/editor/whyvscode)
- [RStudio and features](https://rstudio.com/products/rstudio/features/)
- [Discussion of editors](https://forums.fast.ai/t/text-editor-vs-jupyter-notebook-vs-ides/22069)
- [Jupyter against RStudio](https://stackshare.io/stackups/jupyter-vs-rstudio)
- [VS Code against RStudio](https://stackshare.io/stackups/rstudio-vs-visual-studio-code)
- [Starting with R in VS Code](https://renkun.me/2019/12/11/writing-r-in-vscode-a-fresh-start/)
- [VS Code and Atom](https://www.makeuseof.com/tag/visual-studio-code-vs-atom/)